15yrsChange

### 2024-04-05 doc/15yrsChange.Rmd
minor edits

for submission to Anthromes joint special collection in New Phytologist: https://www.newphytologist.org/news/view/332

@@

*(Word limit: 6,500 words)*

@@

- The future of the important temperate forest carbon (C) sink is dependent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood. Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration. 
- We analyzed biomass change and regenerative capacity in sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
- Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens. There was also severe depletion of canopy tree species in the understory in areas with high deer browsing, endangering future canopy regeneration.
- Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem. This indicates that the C sequestration of U.S. Eastern deciduous forests is likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.

@@

*New Phytologist allows 5-8 keywords. Short phrases are acceptable*

@@

The temperate forest biome plays a critical role in the global carbon cycle (C), accounting for almost half of the net global forest C sink [@harris_global_2021], and temperate deciduous forests alone sequester >300 Tg C yr^-1^ [@pugh_role_2019]. Although it is currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020]<!--check refs-->, and current global C models do not represent certain influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
For instance, the impact of nuisance species, is not represented in global C - i.e., the impact of indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, leading to detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
The impact of nuisance species on forest C budgets is also not considered in machine learning or niche models that seek to project future forest distribution and C stocks, nor is it in C offset projects [@wu_uncertainty_2023].
Nuisance species, including non-indigenous insect pests and pathogens and over-abundant herbivores, are dramatically impacting carbon cycling in temperate deciduous forests around the world, and failure to account for them significantly limits understanding of future forest health and C sequestration potential.

@@

Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*). Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species [*e.g.*, oak, *Quercus* spp.; @bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (NZCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

@@

Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure `r loss_gain_net`b). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals. However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure `r mortality_by_genera`). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.

>>>>

<!-- for submission to Anthromes joint special collection in New Phytologist: https://www.newphytologist.org/news/view/332  -->

@@

@@

- The future of the critical temperate forest carbon (C) sink is contigent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood. Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration. 
- We analyzed biomass change and regenerative capacity in varying sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
- Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens. There was also severe depletion of canopy tree species in the understory in areas with high deer browsing, endangering future canopy regeneration.
- Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem. The C sequestration of Eastern US deciduous forests is therefore likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.

@@

@@

The temperate forest biome plays a critical role in the global carbon cycle (C), accounting for almost half of the net global forest C sink [@harris_global_2021], and temperate deciduous forests alone sequester >300 Tg C yr^-1^ [@pugh_role_2019]. Although it is currently a net C sink, the future of the temperate forest biome remains uncertain.
The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012]. 
Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020], and current global C models do not represent certain influential mechanisms [*e.g.*, @fatichi_moving_2014; @clark_continentwide_2021].
For instance, the impact of nuisance species, is not represented in global C - i.e., the impact of indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, leading to detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010]. 
The impact of nuisance species on forest C budgets is also not considered in machine learning or niche models that seek to project future forest distribution and C stocks, nor is it in C offset projects [@wu_uncertainty_2023].
Nuisance species, including non-indigenous insect pests and pathogens and over-abundant herbivores, are dramatically impacting carbon cycling in temperate deciduous forests around the world, and failure to account for them significantly limits understanding of future forest health and C sequestration potential.

@@

Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*). Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000]. White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a]. Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017). The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].

@@

The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W). The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]. The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

@@

Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure `r loss_gain_net`b). The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens. Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals. However, in the most recent period from 2018 to 2023, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^. Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first to most recent interval. Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.

@@

Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species. The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak(*Quercus velutina*) (Figure `r mortality_by_genera`). *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and loss peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020. *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^. The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^). Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.
